---
title: "Generate Fusion Summary Files"
output: html_notebook
author: Daniel Miller (D3b), Jaclyn Taroni (CCDL), Jo Lynne Rokita (D3b)
date: 2020, 2023
params:
  ci_run:
    label: "1/0 to run in CI"
    value: 1
    input: integer
editor_options: 
  chunk_output_type: inline
---

Generate fusion files specifically for consumption by molecular subtyping analyses

## Set up

### Libraries and functions

```{r}
library(tidyverse)
library(annoFuseData)
```

```{r}
#' Generate filtered fusion frame
#' @param df Unfiltered fusion data frame
#' @param bioid List of biospecimen IDs
#' @param fuses List of explicit fusion names
#' @param genes List of gene names
#' @return the filtered fusion data frame
filterFusion <- function(df, bioid, fuses, genes) {
  if (!missing(bioid)) {
    df <- dplyr::filter(df, Sample %in% bioid)
  }
  if (!missing(fuses) & !missing(genes)) {
    df <- dplyr::filter(df, FusionName %in% fuses |
                   Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  } else if (!missing(fuses)) {
    df <- dplyr::filter(df, FusionName %in% fuses)
  } else if (!missing(genes)) {
    df <- dplyr::filter(df,
                 Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  }
  return(df %>% 
           dplyr::select(Sample, FusionName, Fusion_Type, Gene1A, Gene1B, Gene2A, Gene2B, Gene1A_anno, Gene1B_anno, reciprocal_exists, DomainRetainedGene1A, DomainRetainedGene1B) %>%
           distinct())
}

#' Generate matrix with fusion counts
#' @param fuseDF Filtered fusion data frame
#' @param bioid List of biospecimen IDs that should be included in final table
#' @return Data frame that contains fusion counts
prepareOutput <- function(fuseDF, bioid) {
  fuseDF %>% 
    # some fusions have in-frame and frameshift fusion calls for a sample
    # this will make unique fusionName and Sample dataset to get 1/0 values
    distinct(Sample, FusionName) %>%
    mutate(Count = 1) %>%
    pivot_wider(names_from = FusionName, values_from = Count, values_fill = list(Count = 0)) %>%
    right_join(data.frame(Sample = specimensUnion)) %>%
    replace(is.na(.), 0) %>%
    rename(Kids_First_Biospecimen_ID = Sample)
}
```

### Read in data

```{r}
dataDir <- file.path("..", "..", "data")
fusDir <- file.path("..", "..", "analyses", "fusion_filtering", "results")
annotDir <- file.path("..", "..", "analyses", "fusion_filtering", "references")
#' Annotation file to be used for identifying kinase genes
annot <- read.delim(system.file("extdata", "genelistreference.txt", package = "annoFuseData"))

#' The putative oncogenic fusion file is what we'll use to check for the #' presence or absence of the fusions.
#' However, some biospecimens are not represented in this filtered, prioritized
#' file but have been assayed - we will gather those from the histology file
hist <- read_tsv(file.path(dataDir, "histologies-base.tsv"))

# This needs to come from the data directory when running in CI
if ( params$ci_run == 1 ){
putativeOncogenicDF <- 
  read_tsv(file.path(fusDir, "fusion-putative-oncogenic.tsv"))
} else{
putativeOncogenicDF <- 
  read_tsv(file.path(dataDir, "fusion-putative-oncogenic.tsv"))
}


```

### Read in DGD and arrange as fusion filtering df 
```{r}
# read in DGD data 
dgd_fusion <- readr::read_tsv(file.path(dataDir, "fusion-dgd.tsv.gz"))

# combine the two dataframes
putativeOncogenicDF <- bind_rows(putativeOncogenicDF, dgd_fusion)
```

### Output

```{r}
resultsDir <- "results"
if (!dir.exists(resultsDir)) {
  dir.create(resultsDir)
}
ependFile <- file.path(resultsDir, "fusion_summary_ependymoma_foi.tsv")
embryFile <- file.path(resultsDir, "fusion_summary_embryonal_foi.tsv")
ewingsFile <- file.path(resultsDir, "fusion_summary_ewings_foi.tsv")
lgg_hggFile <- file.path(resultsDir, "fusion_summary_lgg_hgg_foi.tsv")
```

## Fusions and genes of interest

Taken from [`AlexsLemonade/OpenPBTA-analysis#245`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/245), [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251), [`AlexsLemonade/OpenPBTA-analysis#623`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/623) respectively, and [`AlexsLemonade/OpenPBTA-analysis#808`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/808) 

```{r}
#' **Filters**
#'
#' *Fusions Filters*
#' 1: Exact match a list of fusions common in Ependymoma tumors
ependFuses <- c(
  "ZFTA--MAML2",
  "ZFTA--RELA",
  "PTEN--TAS2R1",
  "YAP1--FAM118B",
  "YAP1--MAMLD1",
  "YAP1--MAML2"
)
ependGenes <- c(
  "ZFTA"
)
#' 2: Exact match a list of fusions common in Embryonal tumors
#' as well as fusions containing a particular gene with any other gene
embryFuses <- c(
  "CIC--NUTM1",
  "MN1--BEND2",
  "MN1--CXXC5",
  "EWSR1--PLAGL1"
)
embryGenes <- c(
  "FOXR2",
  "MN1",
  "TTYH1"
)

#' 3: Exact match a list of fusions which are hallmark fusions in Ewings Sarcoma
ewingsFuses<- c(
  "EWSR1--FLI1", 
  "EWSR1--ERG",
  "FUS--ERG"
)

#' 4: Exact match a list of fusions common in low-grade glioma and high-grade glioma tumors
#' as well as fusions containing a particular gene with any other gene
lggFuses <- c(
  "KIAA1549--BRAF",
  "FGFR1--TACC1",
  "MYB--QKI"
)
lgghggGenes <- c(
  "BRAF",
  "ALK",
  "ROS1",
  "NTRK1",
  "NTRK2",
  "NTRK3",
  "PDGFRA",
  "FGFR2",
  "FGFR1",
  "MYB",
  "MYBL1",
  "RAF1", 
  "MET"
)

```

### Filter putative oncogenic fusions list

```{r}
allFuseEpend <- filterFusion(df = putativeOncogenicDF,
                             fuses = ependFuses,
                             genes = ependGenes)
allFuseEmbry <- filterFusion(df = putativeOncogenicDF,
                             fuses = embryFuses,
                             genes = embryGenes)
allFuseEwing <- filterFusion(df = putativeOncogenicDF,
                             fuses = ewingsFuses)
allFuseLGGHGG <- filterFusion(df = putativeOncogenicDF,
                             fuses = lggFuses,
                             genes = lgghggGenes)

```

Get the biospecimen IDs that have had either RNA-Seq or Targeted fusion panel performed.
We use the putative oncogenic file here, therefore any sample that was profiled but does not have a fusion that is relevant to the subtyping tickets is not _missing_ but instead has no evidence of the relevant fusions.

```{r}
# gather and combine
specimensUnion <- hist %>%
  filter((experimental_strategy == "RNA-Seq" | grepl("DGD_FUSIP", aliquot_id)),
         # remove TCGA/GTEx
         !cohort %in% c("TCGA", "GTEx")) %>%
  pull(Kids_First_Biospecimen_ID)
```

#### Write non-MB, non-ATRT embryonal fusions to file

```{r}
allFuseEmbry <- allFuseEmbry %>%
  prepareOutput(specimensUnion)

# Which fusions of interest are not present in any samples?
missingEmbryFusion <- setdiff(embryFuses, colnames(allFuseEmbry)[-1])
# For the fusions that are not present, fill those columns with 0
allFuseEmbry[, missingEmbryFusion] <- 0

# Write to file
allFuseEmbry %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  write_tsv(embryFile)
```

#### Write ependymoma fusions to file

```{r}
allFuseEpend <- allFuseEpend %>%
  prepareOutput(specimensUnion) 

# Which fusions of interest are not present in any samples?
missingEpendFusion <- setdiff(ependFuses, colnames(allFuseEpend)[-1])
# For the fusions that are not present, fill those columns with 0
allFuseEpend[, missingEpendFusion] <- 0

allFuseEpend %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  write_tsv(ependFile)
```


#### Write ewings sarcoma fusions to file

```{r}
allFuseEwing <- allFuseEwing %>%
  prepareOutput(specimensUnion)

# Which fusions of interest are not present in any samples?
missingEwingFusion <- setdiff(ewingsFuses, colnames(allFuseEwing)[-1])
# For the fusions that are not present, fill those columns with 0
allFuseEwing[, missingEwingFusion] <- 0

allFuseEwing %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  write_tsv(ewingsFile)
```

#### Perform selection for LGG and HGG fusions
First pull the fusions or genes from the goi list which are not kinases for the final output file, since these will not need further interrogation.
```{r}
# Which genes/fusions are not kinases, but in the list?
# Separate fusions into genes, combine with gene list, check for not kinase
lggFuses_df <- as.data.frame(lggFuses) %>%
  separate(lggFuses, into = c("Gene1A", "Gene1B"), remove = F)

kinases <- annot %>%
  filter(type == "Kinase") %>%
  pull(Gene_Symbol)

lggFuses_df <- lggFuses_df %>% 
  dplyr::mutate(Gene1A_anno = case_when(Gene1A %in% kinases ~ "Kinase", TRUE ~ "Non-Kinase"), 
                Gene1B_anno = case_when(Gene1B %in% kinases ~ "Kinase", TRUE ~ "Non-Kinase"))

# Only pull fusions that do not contain kinase genes, as ones with kinases will be dealt with separately later
nonkinase_lggFuses <- lggFuses_df %>%
  filter(Gene1A_anno == "Non-Kinase" & Gene1B_anno == "Non-Kinase") %>%
  pull(lggFuses)

# Identify non-kinase genes in LGG goi list
nonkinase_lgghggGenes <- setdiff(lgghggGenes, kinases)

# Pull LGAT non-kinase fusions
nonkinaseLGG <- filterFusion(df = putativeOncogenicDF,
                             fuses = nonkinase_lggFuses,
                             genes = nonkinase_lgghggGenes) %>%
  distinct()
```

Next, collect fusions which contain 3' kinases which are in-frame and retain the kinase domain. 
Keep these for the final output file.
```{r}
three_prime_kinase_inframe <- allFuseLGGHGG %>%
  filter(grepl("Kinase", Gene1B_anno) & Fusion_Type == "in-frame" & DomainRetainedGene1B == "Yes") %>%
  select(Sample, FusionName, Gene1A, Gene1B) %>%
  distinct()

# Are there any fusions that are in-frame, but do not retain the kinase domain? Do they have in-frame fusions retaining the kinase domain in the same fusion?
three_prime_kinase_outframe <- allFuseLGGHGG %>%
  filter(grepl("Kinase", Gene1B_anno) & Fusion_Type == "in-frame" & DomainRetainedGene1B == "No") %>%
  select(Sample, FusionName) %>%
  distinct()

three_prime_kinase_outframe
```

Let's look at these just to be sure the results are as expected.
```{r}
# `BS_KE56MMY0 ARHGEF2--NTRK1` one does and will be captured in the `three_prime_kinase_inframe` list, but `BS_B1C6GZ84 CHIC2--PDGFRA2` does not retain the kinase domain, so we do not want to add it.
intersect(three_prime_kinase_outframe[,c("Sample", "FusionName")], three_prime_kinase_inframe[,c("Sample", "FusionName")])
```

```{r}
# Which fusions are not in-frame?
three_prime_kinase_outframe <- allFuseLGGHGG %>%
  filter(grepl("Kinase", Gene1B_anno) & Fusion_Type != "in-frame") %>%
  distinct()
# Check they don't retain the kinase domain. They do not, so leave out.
table(three_prime_kinase_outframe$DomainRetainedGene1B)
```

Next, filter all fusions for 5' kinase fusions which have reciprocal fusions and retain the kinase domain.
Keep these for the final output file.
```{r}
# Keep those with kinase domain retained and fusion in-frame - keep this list
five_prime_domain_intact <- allFuseLGGHGG %>%
  filter(grepl("Kinase", Gene1A_anno) & DomainRetainedGene1A == "Yes" & Fusion_Type == "in-frame") %>%
  select(Sample, FusionName) %>%
  distinct()
```

Next, filter all fusions for 5' kinase fusions which have lost the kinase domain for reciprocal fusions which have a kinase domain intact and are in-frame.
Keep these for the final output file.
```{r}
# First, get fusions with 5' kinases which lost the kinase domain and have a reciprocal, then add the reciprocal fusion.
five_prime_domain_lost <- allFuseLGGHGG %>%
  filter(grepl("Kinase", Gene1A_anno) & reciprocal_exists == "TRUE") %>%
  select(Sample, FusionName, Gene1A, Gene1B, Fusion_Type, DomainRetainedGene1A) %>%
  filter(DomainRetainedGene1A == "No") %>%
  distinct() %>%
  mutate(three_prime_kinase = paste(Gene1B, Gene1A, sep ="--"))

# Then, select the in-frame reciprocals which have the kinase domain in tact. Retain 5' kinase fusion information and update 3' fusion column name to FusionName for merging with allFuseLGGHGG.
five_prime_kinase_keep <- five_prime_domain_lost %>% 
  select(Sample, five_prime_kinase = FusionName, FusionName = three_prime_kinase) %>%
  left_join(allFuseLGGHGG, by = c("Sample", "FusionName")) %>%
  filter(Fusion_Type == "in-frame" & DomainRetainedGene1B == "Yes") %>%
  select(Sample, FusionName = five_prime_kinase) %>%
  distinct()
```

Rbind lists for final table of LGG and HGG fusions of interest
```{r}
# Rbind lists for final table of LGAT fusions of interest
subsetFuseLGGHGG <- bind_rows(nonkinaseLGG,
                            three_prime_kinase_inframe, 
                            five_prime_domain_intact, 
                            five_prime_kinase_keep) %>%
  select(Sample, FusionName) %>%
  distinct()
```

#### Write LGG and HGG fusions to file

```{r}
subsetFuseLGGHGG <- subsetFuseLGGHGG %>%
  prepareOutput(specimensUnion)

# Which fusions of interest are not present in any samples?
missingLGGFusion <- setdiff(lggFuses, colnames(subsetFuseLGGHGG)[-1])
# For the fusions that are not present, fill those columns with 0
subsetFuseLGGHGG[, missingLGGFusion] <- 0

subsetFuseLGGHGG %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  write_tsv(lgg_hggFile)
```

## Session Info

```{r}
sessionInfo()
```
